---
title: "Spacial Data Science"
format: html
toc: true
---

Based on Spatial Data Science book <https://r-spatial.org/book/>

# Getting started

```{r}
library(tidyverse)
library(sf)
system.file("gpkg/nc.gpkg", package="sf") |>
    read_sf() -> nc
nc.32119 <- st_transform(nc, 'EPSG:32119')
nc.32119 |>
    select(BIR74) |>
    plot(graticule = TRUE, axes = TRUE)
```

```{r}
year_labels <- c("SID74" = "1974 - 1978", "SID79" = "1979 - 1984")
nc.32119 |> select(SID74, SID79) |>
    pivot_longer(starts_with("SID")) -> nc_longer
ggplot() + geom_sf(data = nc_longer, aes(fill = value), linewidth = 0.4) + 
  facet_wrap(~ name, ncol = 1, labeller = labeller(name = year_labels)) +
  scale_y_continuous(breaks = 34:36) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  theme(panel.grid.major = element_line(color = "white"))
```
```{r}
library(mapview) |> suppressPackageStartupMessages()
mapviewOptions(fgb = FALSE)
nc.32119 |> mapview(zcol = "BIR74", legend = TRUE, col.regions = sf.colors)
```

# Coordinates

A reference system = a datum and measurement unit (scale)
A datum = a fixed origin of  location (position) and time

Two-dimensional coordinates:
  -     Cartesian - (x,y)
  -     Polar - (promień, kąt)
  
Three-dimensional:
  -     Cartesian geocentric coordinates (x,y,z)
  -     Spherical/elipsoidal coordinates (promień, kąt (długość), kąt (szerokość) )
  
# CRS

A geodetic datum - datup describin the relationship of 2-d or 3-d coordinate system to the earth
coordinate reference system - a coordinate system that is related to an object by datum. For geodetic and vertical datums the object will be Earth

Coordinate conversion
- recomputing the projections withouth changing datum
- is lossless and invertible

Coordinate transformation
- recomputing coordinates in a new datum
- is approximate

# Geometries


